---
title: "Lichess.org"
description: |
  Analysis of chess game data from TidyTuesday
author: Nikhil Shah
date: February 19, 2025
format: html
---

This analysis uses data from the [October 1, 2024 TidyTuesday](https://github.com/rfordatascience/tidytuesday/tree/main/data/2024/2024-10-01).

This is a dataset of 20,058 online chess games. The data was collected and cleaned from Lichess.org by [Kaggle/Mitchell J.](https://www.kaggle.com/datasets/datasnaek/chess/data) It includes extensive data on each game including each move used, both players' usernames and ratings, and the time of the game.

```{r}
#| label: Data Import
#| echo: false
#| include: false

library(tidyverse)

chess <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-10-01/chess.csv')
```

```{r}
#| label: Data Wrangling and Barplot

# adds a column that isolates the opening move of each game, and orders it by how common each move is
chess <- chess |>
  mutate(
    opening_move = ifelse(
      str_sub(moves, 1, 1)=='N', 
      str_sub(moves, 1, 3), 
      str_sub(moves, 1, 2)
      )
    ) |>
  mutate(
    opening_move = fct_infreq(opening_move)
    )

# creates a bar plot that shows the frequency of opening moves
chess |>
  ggplot() +
  geom_bar(aes(x = opening_move)) +
  labs(
    x = 'Opening Move',
    y = 'Number of Games',
    title = 'Frequency of opening moves for white on Lichess.org',
    alt = 'Bar chart showing the frequency of opening moves for white on Lichess.org, ordered by frequency. e4, d4, Nf3, c4 are the highest, the rest are largely negligible.'
  )
  
```

This bar plot shows the frequency of initial moves in the dataset. e4 and d4 make up almost all of them. While knight to f3 is the third most popular, almost none of the games use knight to a3.

```{r}
#| label: Extra Data
#| echo: false

chess |>
  group_by(opening_move) |>
  summarize(count = n()) |>
  mutate(proportion = count/20058*100)
```
