---
title: "National Park Species"
author: "Nikhil Shah"
subtitle: "May 13, 2025"
format:
  revealjs:
    scrollable: true
    slide-number: true
    show-slide-number: all
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Project 1 (Data Visualization)

- TidyTuesday
- 10/8/24, National Park Species
- Dataset: every species in every U.S. national park

## Initial Project

- Very simple and mostly meaningless chart of how many species of one specific animal family are in various national parks
```{r}
#| warning: false
#| echo: false

library(tidyverse)
library(shiny)

species <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-10-08/most_visited_nps_species_data.csv')

species <- species |>
  mutate(ParkName = str_replace(ParkName, ' National Park', '')) |>
  mutate(ParkName = fct_infreq(ParkName))
```
```{r}
#| eval: false
species |>
  filter(Family == 'Mustelidae') |>
  ggplot(aes(y = ParkName)) +
  geom_bar() +
  labs(
    y = 'National Park',
    x = '# Species',
    title = 'Distinct species in family Mustelidae in 15 national parks'
  )
```

## Plot
```{r}
#| echo: false
species |>
  filter(Family == 'Mustelidae') |>
  ggplot(aes(y = ParkName)) +
  geom_bar() +
  labs(
    y = 'National Park',
    x = '# Species',
    title = 'Distinct species in family Mustelidae in 15 national parks'
  )
```

## Updated Project

- Interactive Shiny tool to control which mammal taxon is displayed by the plot
- Also not very useful but much more fun and informative
- Shows a randomly-sampled common species name from the selected family to make it more understandable
- [Link to code](https://nikhilshah7.github.io/dataviz3.html)
- [Link to app](https://nikhilshah.shinyapps.io/species/)