---
title: "Reverse 3-0s"
description: |
  Simulating best-of-5 tournament matches to estimate the chances of a reverse 3-0
author: Nikhil Shah
date: April 2, 2025
format: html
---

In a best-of-5 set, a reverse 3-0 occurs when one side loses the first two games, and then wins the next three to win a set. It's an uncommon and extremely exciting (or heartbreaking) phenomenon, and it can be pretty difficult to predict when it might occur. I want to estimate how frequently this happens when a set starts 2-0, depending on the relative strength of both sides.

This needs some assumptions. Firstly, the chance of one player or team to win a game heavily depends on relative skill, the volatility of the game, and maybe luck. Without looking at data from a specific game, we can create chances that seem to make sense.

We also have to look at players/teams of a somewhat comparable skill. If you were watching a best-of-5 tournament match, and you saw one side go down 0-2, you would only wonder if they could win the match if they were capable of making it competitive. In a 1024-entrant bracket, odds are, the 1st seed will beat the 1024th seed, and the 512th seed, and the 256th seed handily in a 3-0. While those could technically be a reverse 3-0 after the first two games, it's very unlikely and nobody would expect it to happen. We want to look at matches where some competitiveness is expected.

We can use a 16-player tournament, and assume that, in a vacuum, the odds of one side winning a game are the the ratio of the two teams' seeds. For instance, seed 1 has a  1:2 chance of beating seed 2 (67%), seed 16 has a 16:1 chance of beating seed 1 (6%), and seed 15 has 15:16 odds of beating seed 16 (48%). Each of these matches would be competitive, with a significant degree of variability, which makes it interesting.

Let's create a hypothetical match between a random pairing of two seeds from 1-16. We can take one side's odds of winning a game to the power of three and determine the chance of them winning three games in a row. However, since we want to know the chance of a reverse sweep rather than just a sweep, we need to account for the fact that the player or team that goes up 2-0 might have some additional advantage specifically related to that opponent: experience in the matchup, knowledge of habits or strategies, a favorable difference in skill sets, etc. Therefore, the other side's chance of winning a game is most likely lower than initially assumed.

How can we account for this? If you flip a coin twice, 25% of the time, it'll land on tails twice by pure random chance. In that spirit, let's say that when a set starts 2-0, 75% of the time it's because of a difference in skill or preparation, and 25% of the time it happened just because of variability with no real pattern beneath it. We can take the losing side's initial estimated chance of winning, and reduce it by 25% to account for this factor. Based on my general experience with spectating tournaments, this is a somewhat accurate measure.

```{r}
#| warning: false

library(tidyverse)
library(purrr)
```


```{r}
# Precursor for the upset_factor() function
expected_placement <- function(x) {
  if (x >= 1 & x <= 4) {
    order <- x
  }
  else if (x == 5 | x == 6) {
    order <- 5
  }
  else if (x ==7 | x == 8) {
    order <- 6
  }
  else if (x >= 9 & x <= 12) {
    order <- 7
  }
  else if (x >= 13 & x <= 16) {
    order <- 8
  }
}

# This scales the difference in seeds according to the difference in their expected placements in a double elimination bracket. Seed 1 vs Seed 4 is a similar difference to Seed 5 vs Seed 15. 
upset_factor <- function(x, y) {
  upset <- expected_placement(x) - expected_placement(y)
}

# Randomizes seeds of two opponents, determines the odds of the second team winning games after going down 0-2, finds the seed differential from the previous functions.
reverse_sweep_chance <- function() {
  first_two_winner <- sample(1:16, 1)
  reverse_sweep_candidate <- sample(1:16, 1)
  win_odds <- (first_two_winner / (first_two_winner + reverse_sweep_candidate) * 0.75)
  did_they_win <- (runif(1) < win_odds * win_odds * win_odds)
  seed_comparison <- upset_factor(reverse_sweep_candidate, first_two_winner)
  return(data.frame(reverse_sweep = did_they_win, upset_factor = seed_comparison))
}

# Repeats the test
data <- map(1:1000, ~ reverse_sweep_chance()) |>
  list_rbind() 

# Finds the overall total of reverse sweeps that happened
data |>
  summarize(chance = mean(reverse_sweep))

# Plots the data
data |>
  ggplot(aes(x = upset_factor, fill = reverse_sweep)) +
  geom_bar(position = 'fill') +
  labs(
    x = 'Upset Factor',
    y = 'Reverse 3-0 Chance',
    fill = 'Did It Happen?',
    title = 'Reverse Sweep Chance Depending on Seed Differential',
    alt = 'Bar chart showing the chance of a reverse sweep for a range of seed differentials, based on a simulation of 1000 trials. Probability decreases linearly, and ranges from about 40% for -8 to 0% for 4 and above.'
  )

```

I think this ended up being a very good estimation of the chances of this happening across different seeds. While relatively equally-skilled opponents can rarely make a reverse sweep, when the opponent that's up 2-0 is marginally higher seeded it almost never happens. Meanwhile, when the lower seeded player goes up early, they still stand a very significant chance of getting reverse swept. This simulation largely aligns with my experience with competitive games and sports. 

This simulation helps to illustrate how likely (or unlikely) a major comeback is, even with a high degree of volatility. Conversely, it shows how an early match lead in a best of 5 does not remotely guarantee victory against a better opponent. While the exact numbers definitely could be adjusted for more accurate results across different games, I think this does a good job of representing the probability of a reverse sweep depending on the strength of one's opponent. Despite the fact that the odds of winning are completely made up, I do think the simulation lends credibility to the idea that the likelihood of this phenomenon can be extremely variable depending on the relative skill levels of two opponents.
