---
title: "SQL: Hungry Cop Effect"
description: |
  Is there a pattern in the outcome of traffic stops based on whether they're after lunch?
author: Nikhil Shah
date: April 22, 2025
format: html
---

## Introduction

It's been observed that judges are less likely to grant lenient sentences or parole when they are hungry. (i.e. before their lunch break) (Danziger et al., 2011). This has become known as the "hungry judge effect".

We can use data from the Stanford Open Policing Project (Pierson et al., 2020) to observe the rate at which traffic stops result in citations, and whether there's a meaningful difference in this rate before and after noon (which, presumably, is roughly when cops would eat lunch and be the hungriest). Washington, Texas, and New Jersey all give data for citations given and time of day. 

If we assume that a standard lunch hour is 12-1pm, we can consider the prior hour (11-12) "before lunch" and the next hour (1-2) "after lunch". While traffic stops obviously happen outside of those hours, it doesn't make sense to evaluate the effect of a lunch break on something happening in the middle of the night, or even at 9am or 5pm.

## Data

### Libraries

```{r}
#| label: libraries
#| warning: false

library(RMariaDB)
library(tidyverse)
```


### SQL Connection

```{r}
#| label: connection

con_traffic <- DBI::dbConnect(
  RMariaDB::MariaDB(),
  dbname = "traffic",
  host = Sys.getenv("TRAFFIC_HOST"),
  user = Sys.getenv("TRAFFIC_USER"),
  password = Sys.getenv("TRAFFIC_PWD")
)
```

### Query

```{sql}
#| label: query
#| connection: con_traffic
#| output.var: 'citations_and_warnings'

SELECT 'WA' AS state,
       CASE
        WHEN time >= '11:00:00' AND time <= '12:00:00' THEN 'morning'
        WHEN time >= '13:00:00' AND time <= '14:00:00' THEN 'afternoon'
       END AS day_segment,
       AVG(citation_issued) AS citation_rate,
       AVG(warning_issued) AS warning_rate,
       COUNT(*) AS total_stops
FROM wa_statewide_2020_04_01
WHERE time >= '11:00:00' AND time <= '12:00:00' OR time >= '13:00:00' AND time <= '14:00:00'
GROUP BY day_segment

UNION ALL

SELECT 'TX' AS state,
       CASE
        WHEN time >= '11:00:00' AND time <= '12:00:00' THEN 'morning'
        WHEN time >= '13:00:00' AND time <= '14:00:00' THEN 'afternoon'
       END AS day_segment,
       AVG(citation_issued) AS citation_rate,
       AVG(warning_issued) AS warning_rate,
       COUNT(*) AS total_stops
FROM tx_statewide_2020_04_01
WHERE time >= '11:00:00' AND time <= '12:00:00' OR time >= '13:00:00' AND time <= '14:00:00'
GROUP BY day_segment

UNION ALL

SELECT 'NJ' AS state,
       CASE
        WHEN time >= '11:00:00' AND time <= '12:00:00' THEN 'morning'
        WHEN time >= '13:00:00' AND time <= '14:00:00' THEN 'afternoon'
       END AS day_segment,
       AVG(citation_issued) AS citation_rate,
       AVG(warning_issued) AS warning_rate,
       COUNT(*) AS total_stops
FROM nj_statewide_2020_04_01
WHERE time >= '11:00:00' AND time <= '12:00:00' OR time >= '13:00:00' AND time <= '14:00:00'
GROUP BY day_segment;
```

This takes the overall data for a state, filters out times outside of the normal workday, adds a column that specifies what part of the day a stop occurred at based on the time, and splits the table into two parts for each part of the day. It averages the column "citation_issued", which is 1s for yes and 0s for no, to get the rate at which citations were issued at each part of the day. It combines this output for three different states into one table.

```{r}
#| label: table and disconnect
dbDisconnect(con_traffic)

citations_and_warnings
```

The final data from the SQL query shows one row for each pairing of state and part of the day, and the proportion of traffic stops resulting in a citation or a warning in each. 

## Visualization

```{r}
#| label: plots

citations_and_warnings |>
  ggplot() +
  geom_col(
    aes(x = state, y = citation_rate, fill = day_segment),
    position = 'dodge'
  ) +
  labs(
    x = 'State',
    fill = 'Part of Day',
    y = 'Citation Rate',
    title = 'Frequency of Citations from Traffic Stops at Different Times of Day',
    alt = 'Bar chart showing the frequency of citations from traffic stops from 11am to 12pm, and 1pm to 2pm, in New Jersey, Washington, and Texas. In each state, the frequency for the earlier hour is slightly higher. Overall citation rate is much higher in New Jersey.'
  )

citations_and_warnings |>
  ggplot() +
  geom_col(
    aes(x = state, y = warning_rate, fill = day_segment),
    position = 'dodge') +
  labs(
    x = 'State',
    fill = 'Part of Day',
    y = 'Warning Rate',
    title = 'Frequency of Warnings from Traffic Stops at Different Times of Day',
    alt = 'Bar chart showing the frequency of warnings from traffic stops from 11am to 12pm, and 1pm to 2pm, in New Jersey, Washington, and Texas. Bars for each hour are not meaningfully different. Overall warning rate is much higher in Texas.'
  )
```


## Interpretation

The answer seems to be "maybe". While 42% of morning traffic stops in Washington state result in citations, compared to 40% of afternoon ones, this pattern is much weaker in the other two states, with a difference of about half a percentage point in each. The rate at which warnings are given is essentially identical. With a sample size of over a million for Washington and Texas, it is possible that such a small difference is still statistically significant.

Notably, without the time constraint (just observing differences between before and after 12pm), the morning citation rates are considerably higher than the afternoon ones for all three states. I assume this is because "morning" encompasses late-night traffic stops, which evidently result in much more citations.

Overall, I don't think these findings are conclusive. Big states like California and New Jersey do not supply time of citation information, so it may be helpful to collect more data to confirm these findings. I don't want to rule out the possibility that such an effect exists but if it is it's much less blatant than it is when observed in judges.


## References

 * [Danziger, Shal, Jonathan Levav, Liora Avnaim-Pesso (2011). "Extraneous Factors in Judicial Decisions." *Proc. Natl. Acad. Sci. U.S.A.*, 108 (17) 6889-6892.](https://www.pnas.org/doi/full/10.1073/pnas.1018033108)

 * [Pierson, Emma, Camelia Simoiu, Jan Overgoor, Sam Corbett-Davies, Daniel Jenson, Amy Shoemaker, Vignesh Ramachandran, et al. (2020). “A Large-Scale Analysis of Racial Disparities in Police Stops Across the United States.” *Nature Human Behaviour*, 1–10.](https://www.nature.com/articles/s41562-020-0858-1)


